---
title: "Demand Forecasting"
output:
  html_document: default
  pdf_document: default
---

```{r setup , include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### This script uses historical demand data for 17,000 spare parts to forecast short-term demand. 
#### Markov Chain transition probabilities and bootsrapping are used to deal with the intermittent naure of spare parts demand.


```{r , include=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(reshape2)
library(sqldf)
library(foreach)
library(doParallel)
library(tsintermittent)
library(boot)
library(markovchain)

data1 <- read.csv("DemandData.csv")
data1 <- data1[,-c(1)]
data1$RequestDate <- as.Date(data1$RequestDate)
data1$Catalog_ID <- as.character(data1$Catalog_ID)
data1$Demand <- as.numeric(data1$Demand)
```

#### Create new demand variable: zero demand = a,nonzero demand = b 
```{r , include=TRUE}
data1$Demand[is.na(data1$Demand)==TRUE] <- 0
data1$Demand2[data1$Demand==0] <- "a"
data1$Demand2[data1$Demand>0] <- "b"
data1$Demand2[data1$Demand<0] <- "b"
data1$Demand2 <- as.character(data1$Demand2)
```


### Training Data:
#### Back-up and subset to orders before April 2016.
#### We will forecast April 2016 - July 2016 and compare to actuals 
```{r , include=TRUE}
hold <- data1
data1 <- data1[data1$RequestDate < '2016-04-01',]

sixteen <- hold$Catalog_ID[substr(hold$RequestDate,1,4) == '2016' & hold$Catalog_ID %in% c(data1$Catalog_ID)]
data1 <- data1[data1$Catalog_ID %in% c(sixteen),]
sample <- sample(unique(data1$Catalog_ID), 1500)
data1 <- data1[data1$Catalog_ID %in% c(sample),]

# item index for loop
df_list = list(c(unique(data1$Catalog_ID)))
```

#### Introduce multicore parallel processing 
```{r , include=TRUE}
cores <- detectCores()
cl <- makeCluster(cores[1]-4)
registerDoParallel(cl)
```

#### Forecast loop for each item
```{r , include=TRUE}
resultsFinal <- data.frame(
  Catalog_ID = as.integer(),
  RequestDate = as.Date(character()),
  fcast = as.character(), run1000_iter1 = as.numeric())

results1 <- foreach(i = 1:length(unique(data1$Catalog_ID)), .combine = rbind, .packages = 'markovchain') %dopar%  {
  
  data2 <- subset(data1, data1$Catalog_ID %in% df_list[[1]][i])
  
  data2$RequestDate <- as.Date(data2$RequestDate)
  
  # Step 1: fit Markov model #
  mysequence <- data2$Demand2
  myFitM <- markovchainFit(mysequence,confidencelevel = .95,method = "mle")
  mc <- myFitM$estimate
  
  transitionMatrix = mc
  
  # get initial state #
  demandToday = data2$Demand2[data2$RequestDate==max(data2$RequestDate)]
  
  # more accurate with more runs #
  runs <- 1000
  
  # number of iterations - started with only 1, but saw improvement when doing > 1 
  # and taking average #
  demand_draws <- 8
  
  ltd_out <- c()
  ltd_out2 <- c()
  
  # Step 2: forecast demand (0,1) #
  for(run in 1:runs){
    
    
    fcast_periods <- 4
    
    initial_state <- if(demandToday == "b"){c(0,1)}else{c(1,0)}
    names(initial_state) <- c('a','b')
    
    tmat <- as.data.frame(transitionMatrix@transitionMatrix)
    
    # loop breaks for items that have a probability of zero in the transition matrix (ie, their historical demands are #
    # all/almost all zero or nonzero) #
    # solution is to make forecast either all 'a' or all 'b' #
    if (nrow(tmat) == 1 & names(tmat) == 'a' ){
      fcast <- c("a","a","a","a")
      fstates <- matrix(c(1,1,1,1,0,0,0,0),nrow = 2, ncol = 2)
    } else if (!is.na(tmat$b[2]) & (tmat$b[1] == 0 | tmat$b[2] == 0) ){
      fcast <- c("a","a","a","a")
      fstates <- matrix(c(1,1,1,1,0,0,0,0),nrow = 2, ncol = 2)
    } else if (nrow(tmat) == 1 & names(tmat) == 'b' ){
      fcast <- c("b","b","b","b")
      fstates <- matrix(c(0,0,0,0,1,1,1,1),nrow = 2, ncol = 2)
    } else {
      
      fcast <- c()
      fstates <- matrix(nrow = fcast_periods,ncol = nrow(tmat))
      
      
      
      for(x in 1:fcast_periods){
        current_state <- if(x == 1){initial_state}else{current_state}
        
        ttemp <- tmat[names(current_state)[current_state == 1],]
        
        new_state <- sample(names(ttemp),1,prob = ttemp)
        
        current_state <- ifelse(names(ttemp) == new_state,1,0)
        names(current_state) <- names(ttemp)
        fcast[x] <- new_state
        fstates[x,] <- current_state
      }
    } 
    
    
    
    resultsTemp <- data.frame(df_list[[1]][i], RequestDate=seq(as.Date("2016/04/1"), as.Date("2016/07/1"), "month")
                              , fcast)
    
    
    #Step 3: bootstrapping - if fcast is b, assign value from sample of historical nonzero demands # 
    for(dmddrw in 1:demand_draws){
      d <- data2[,'Demand'][data2$Demand != 0] #(only nonzero demands used)#
      resultsTemp$non0bs <- ifelse(resultsTemp$fcast == "a",0,d[sample(length(d),1)])
      
      
      for(x in seq(1,nrow(resultsTemp))){
        jittered <-ifelse(resultsTemp$fcast == "a",0,jitter(resultsTemp[x,'non0bs']))
        resultsTemp[x,'non0bs'] <- jittered[x] 
      }
      
      names(resultsTemp) <- ifelse(names(resultsTemp) == "non0bs",paste0("run",run,"_iter",dmddrw),names(resultsTemp))
    }
    
    # Step 4: sum forecast results over horizon
    # produce a distro of ltd based on the simulations above #
    
    ltd_distro <- colSums(resultsTemp[names(resultsTemp)[grepl("run",names(resultsTemp))]])
    
    ltd1 <- sum(resultsTemp[paste0("run",run,"_iter1")])
    ltd_out[run] <- ltd1
    ltd2 <- sum(ltd_distro) / length(ltd_distro)
    ltd_out2[run] <- ltd2
    print(paste0("Finished run ",run))
  } 
  
  #resultsFinal <- rbind(resultsFinal, resultsTemp)
  resultsTemp
}

stopCluster(cl)

names(results1)[1] <- "Catalog_ID"
names(results1)[3] <- "Demand"
names(results1)[4] <- "FcastDemand"
names(results1)[5] <- "FcastDemand2"
names(results1)[6] <- "FcastDemand3"
names(results1)[7] <- "FcastDemand4"
names(results1)[8] <- "FcastDemand5"
names(results1)[9] <- "FcastDemand6"
names(results1)[10] <- "FcastDemand7"
names(results1)[11] <- "FcastDemand8"


results1$Catalog_ID <- as.character(results1$Catalog_ID)
results1$Demand <- as.character(results1$Demand)
results1$RequestDate <- as.character(results1$RequestDate)

save <- results1
```
#### Validation - compare actual demand to forecast
```{r , include=TRUE}
sample <- unique(results1$Catalog_ID)

actuals <- unique(hold[hold$Catalog_ID %in% c(sample) & substr(hold$RequestDate,1,7) %in% c('2016-04','2016-05','2016-06','2016-07'), names(hold) %in% c("Catalog_ID","RequestDate", "Demand")])
names(actuals)[3] <- "ActualDemand"

actuals$RequestDate <- as.character(actuals$RequestDate)

results <- merge(results1, actuals, by = c("RequestDate","Catalog_ID"), all=TRUE)
```

#### Mean Absoulte Error
```{r , include=TRUE,echo=TRUE}
mean(abs(results$FcastDemand - actuals$ActualDemand))
mean(abs(results$FcastDemand2 - actuals$ActualDemand))
mean(abs(results$FcastDemand3 - actuals$ActualDemand))
```

#### Mean Squared Error
```{r , include=TRUE,echo=TRUE}
mean((results$FcastDemand - results$ActualDemand)^2)
mean((results$FcastDemand2 - results$ActualDemand)^2)
mean((results$FcastDemand3 - results$ActualDemand)^2)
```


#### Actual vs Forecast comparison plot
```{r , include=TRUE,echo=TRUE}

ForecastNew <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                          RequestDate = as.Date(results$RequestDate),
                          Demand = as.numeric(results$FcastDemand),
                          Type = as.character("Forecasted"))
ActualNew <- data.frame(Catalog_ID = as.character(actuals$Catalog_ID),
                        RequestDate = as.Date(actuals$RequestDate),
                        Demand = as.numeric(actuals$ActualDemand),
                        Type = as.character("Actual"))
ForecastNew2 <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                           RequestDate = as.Date(results$RequestDate),
                           Demand = as.numeric(results$FcastDemand2),
                           Type = as.character("Forecasted2"))
ForecastNew3 <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                           RequestDate = as.Date(results$RequestDate),
                           Demand = as.numeric(results$FcastDemand3),
                           Type = as.character("Forecasted3"))
ForecastNew4 <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                           RequestDate = as.Date(results$RequestDate),
                           Demand = as.numeric(results$FcastDemand4),
                           Type = as.character("Forecasted4"))
ForecastNew5 <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                           RequestDate = as.Date(results$RequestDate),
                           Demand = as.numeric(results$FcastDemand5),
                           Type = as.character("Forecasted5"))
ForecastNew6 <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                           RequestDate = as.Date(results$RequestDate),
                           Demand = as.numeric(results$FcastDemand6),
                           Type = as.character("Forecasted6"))
ForecastNew7 <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                           RequestDate = as.Date(results$RequestDate),
                           Demand = as.numeric(results$FcastDemand7),
                           Type = as.character("Forecasted7"))
ForecastNew8 <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                           RequestDate = as.Date(results$RequestDate),
                           Demand = as.numeric(results$FcastDemand8),
                           Type = as.character("Forecasted8"))



ResultsNew <- rbind(ForecastNew, ActualNew,ForecastNew2,ForecastNew3,ForecastNew4,ForecastNew5,ForecastNew6,ForecastNew7,ForecastNew8)

test <- sqldf("select distinct RequestDate, Type, sum(Demand) as Demand
              from ResultsNew
              group by RequestDate, Type")
library(ggplot2)
ggplot(test, aes(RequestDate, Demand,group=Type, color=Type)) +
  geom_line() +
  ggtitle("Actual vs Forecasted Demand: 2016")

```

#### Average demand iterations 
```{r , include=TRUE,echo=TRUE}
mean <- rowMeans(results[4:11])
results$MeanForecast <- mean

Forecast <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                       RequestDate = as.Date(results$RequestDate),
                       Demand = as.numeric(results$MeanForecast),
                       Type = as.character("Forecasted"))
Actual <- data.frame(Catalog_ID = as.character(results$Catalog_ID),
                     RequestDate = as.Date(results$RequestDate),
                     Demand = as.numeric(results$ActualDemand),
                     Type = as.character("Actual"))



Results2 <- rbind(Forecast,Actual)

test2 <- sqldf("select distinct RequestDate, Type, sum(Demand) as Demand
               from Results2
               group by RequestDate, Type")
library(ggplot2)
ggplot(test2, aes(RequestDate, Demand,group=Type, color=Type)) +
  geom_line() +
  ggtitle("Actual vs Forecasted Demand: 2016")
```

#### Save results 
```{r , include=TRUE}
final <- sqldf("select distinct RequestDate, Catalog_ID, Demand, ActualDemand, MeanForecast from results")
final$RequestDate <- as.character(final$RequestDate)
```


